---
title: "Theoretic-spectra-generation-of-SIP-labeled-compound"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Theoretic-spectra-generation-of-SIP-labeled-compound}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10
)
```

```{r, include = FALSE, eval=FALSE}
rmarkdown::render("Theoretic-spectra-generation-of-SIP-labeled-compound.Rmd", output_dir = "../doc/")
```

```{r setup}
library(Aerith)
library(ggplot2)
```

### Monte Carlo method

This algorithm calculates the isotope numbers for Stable Isotope Probing (SIP) based on a given chemical formula. It uses a Monte Carlo simulation approach while maintaining the isotopic fine structure of each isotope.

```{r}
# natural abundance of (M-H)- of glucose
iso1<- cal_isotope_numbers_SIP("C6H11O6")
plotMolecularIsotopes(iso1) + 
  ggtitle(expression(C[6]*H[11]*O[6]^"-"~1.07*"% "*{}^{13}*C))
# adjusted abundance in SIP
iso2<- cal_isotope_numbers_SIP("C6H11O6", num_simulations = 50000, C13 = 0.5)
plotMolecularIsotopes(iso2) + 
  ggtitle(expression(C[6]*H[11]*O[6]^"-"~50*"% "*{}^{13}*C))
```

```{r}
# natural abundance of (M+Na)⁺ of glucose
iso1<- cal_isotope_peaks_fft("C6H12O6Na")
plotMolecularFFTisotopes(iso1) + 
  ggtitle(expression(C[6]*H[12]*O[6]*Na^"+"~1.07*"% "*{}^{13}*C))
# adjusted abundance in SIP
iso2<- cal_isotope_peaks_fft("C6H12O6Na", N_width = 200, min_abundance = 0.001, C13 = 0.5)
plotMolecularFFTisotopes(iso2) + 
  ggtitle(expression(C[6]*H[12]*O[6]*Na^"+"~50*"% "*{}^{13}*C))
```

### FFT-based algorithm

This algorithm calculates the isotope peaks for a given chemical formula using Fast Fourier Transform (FFT). It approximates the delta mass of isotopes is one neutron mass and ignore the fine structure of isotopes.

```{r}
# natural abundance of (M-H)- of glucose
iso1<- cal_isotope_peaks_fft("C6H11O6")
plotMolecularFFTisotopes(iso1) + 
  ggtitle(expression(C[6]*H[11]*O[6]^"-"~1.07*"% "*{}^{13}*C))
# adjusted abundance in SIP
iso2<- cal_isotope_peaks_fft("C6H11O6", N_width = 200, min_abundance = 0.001, C13 = 0.5)
plotMolecularFFTisotopes(iso2) + 
  ggtitle(expression(C[6]*H[11]*O[6]^"-"~50*"% "*{}^{13}*C))
```

```{r}
# natural abundance of (M+Na)⁺ of glucose
iso1<- cal_isotope_peaks_fft("C6H12O6Na")
plotMolecularFFTisotopes(iso1) + 
  ggtitle(expression(C[6]*H[12]*O[6]*Na^"+"~1.07*"% "*{}^{13}*C))
# adjusted abundance in SIP
iso2<- cal_isotope_peaks_fft("C6H12O6Na", N_width = 200, min_abundance = 0.001, C13 = 0.5)
plotMolecularFFTisotopes(iso2) + 
  ggtitle(expression(C[6]*H[12]*O[6]*Na^"+"~50*"% "*{}^{13}*C))
```

```{r session-info}
sessionInfo()
```
